### 2022-03-11 ehoh/ehoh.tex
intro

In first-order logic, the arity of a variable is always 0, and the arity of a
symbol is given by its type signature.
In higher-order logic, variables may have function type and be applied, and
symbols can be applied to fewer arguments than specified by their type
signatures. A natural representation of \lfhol{} terms as tagged unions
would distinguish between variables~$x$, symbols~$\cst{f}$, and binary
applications $s \; t.$ However, this scheme suffers from memory
fragmentation and linear-time access, as with the representation of types,
affecting performance on purely or mostly first-order problems. Instead, we
propose a flattened representation, as a generalization of E's existing data
structures: Allow arguments to variables, for symbols let \verb|arity| be
the number of \begin{conf}\emph{actual} arguments\end{conf}\begin{rep}actual
arguments, %as opposed to the declared arity,
and rename the field \verb|num_args|\end{rep}. This representation, often called
``spine notation,'' % \cite{cervesato-pfenning-2003},
is isomorphic to the standard definition of higher-order terms with binary
application. It is employed in various higher-order reasoning systems,
including Leo-III \cite{sb-21-leo3} and
Zipperposition \cite{bbtvw-21-sup-lam}.

@@

\looseness=-1
To ascertain the effectiveness of our approach, we evaluated Ehoh against
E used on applicative encodings of problems (denoted by \appE{}). For reference,
we also evaluated the following versions of higher-order provers that competed in the
THF division of the 2019 edition of CASC \cite{gs-19-casc27}:\
CVC4
1.8 prerelease \cite{brotb-19-ho-smt}, Leo-III 1.4 \cite{sb-21-leo3},
Satallax 3.4 \cite{cb-12-satallax}, Vampire 4.4 \cite{br-19-restricted-unif},
and Zipperposition 1.6 \cite{bbtvw-21-sup-lam}.
Like at CASC, we used different versions of Vampire for first-order
and higher-order problems. Similarly, Zipperposition does not use E as backend when
it is run on first-order problems
and uses different heuristics on first-~and higher-order problems.
The genuine higher-order provers have
the unfair advantage that they can instantiate higher-order
variables with $\lambda$-terms. % (or equivalent SK-style combinator terms).
Thus, some formulas that are
provable by these systems may be nontheorems for \appE{} and Ehoh\confrep{}{, or
they may require tedious reasoning about $\lambda$-lifted functions or {\sf SKBCI}-style
combinators}.
An example is the conjecture $\exists f.\>\forall x\> y.\; f\>x\;y \eq
\cst{g}\;y\;x$, whose proof requires taking $\lambda x\> y.\; \cst{g}\;y\;x$
as the witness for~$f$.

@@

  \begin{sloppypar}
  Traditionally, most higher-order provers were designed from the ground up
  to target higher-order logic. Two
  exceptions are Otter-$\lambda$ by Beeson \cite{mb-04-lam-logic} and
  Zipperposition by Cruanes et al.\ %, Vukmirovi\'c, and colleagues
  \cite{sc-15-simon-phd,bbtvw-21-sup-lam}. \hbox{Otter-$\lambda$}
  adds $\lambda$-terms and second-order unification to
  the superposition prover Otter\begin{rep} \cite{mcc-03-otter}\end{rep}.
  % The approach is pragmatic. , with little emphasis on completeness.
  Zipperposition, also based on superposition, was extended
  to Bool\-e\-an-free higher-order logic by Bentkamp et al.\
  \cite{bbtvw-21-sup-lam}. Its performance is a far cry from E's, but
  it is easier to modify.
  Vukmirovi\'c et al.\ also used it to test and evaluate higher-order unification
  procedures \cite{vbn-21-unif} and Boolean reasoning
  \cite{vn-20-bools}.
  Zipperposition now includes Ehoh as a backend in a cooperative
  architecture.
  Finally, there is work by the developers of Vampire
  % br-20-full-sup-w-combs
  \cite{br-19-restricted-unif} and of the SMT solvers CVC4 and veriT
  \cite{brotb-19-ho-smt} to extend their provers to higher-order logic.
  \end{sloppypar}

>>>>

In first-order logic, the arity of a variable is always 0, and the arity of a
symbol is given by its type signature.
In higher-order logic, variables may have function type and be applied, and
symbols can be applied to fewer arguments than specified by their type
signatures. A natural representation of \lfhol{} terms as tagged unions
would distinguish between variables~$x$, symbols~$\cst{f}$, and binary
applications $s \; t.$ However, this scheme suffers from memory
fragmentation and linear-time access, as with the representation of types,
affecting performance on purely or mostly first-order problems. Instead, we
propose a flattened representation, as a generalization of E's existing data
structures: Allow arguments to variables, for symbols let \verb|arity| be
the number of \begin{conf}\emph{actual} arguments\end{conf}\begin{rep}actual
arguments, %as opposed to the declared arity,
and rename the field \verb|num_args|\end{rep}. This representation, often called
``spine notation,'' % \cite{cervesato-pfenning-2003},
is isomorphic to the standard definition of higher-order terms with binary
application. It is employed in various higher-order reasoning systems,
including Leo-III \cite{sb-21-leo3} and
Zipperposition \cite{sc-15-simon-phd,sc-supind-17}.

@@

\looseness=-1
To ascertain the effectiveness of our approach, we evaluated Ehoh against
E used on applicative encodings of problems (denoted by \appE{}). For reference,
we also evaluated the following versions of higher-order provers that competed in the
THF division of the 2019 edition of CASC \cite{gs-19-casc27}:\
CVC4
1.8 prerelease \cite{brotb-19-ho-smt}, Leo-III 1.4 \cite{sb-21-leo3},
Satallax 3.4 \cite{cb-12-satallax}, Vampire 4.4 \cite{br-19-restricted-unif},
and Zipperposition 1.6 \cite{sc-supind-17,sc-15-simon-phd}.
Like at CASC, we used different versions of Vampire for first-order
and higher-order problems. Similarly, Zipperposition does not use E as backend when
it is run on first-order problems
and uses different heuristics on first-~and higher-order problems.
The genuine higher-order provers have
the unfair advantage that they can instantiate higher-order
variables with $\lambda$-terms. % (or equivalent SK-style combinator terms).
Thus, some formulas that are
provable by these systems may be nontheorems for \appE{} and Ehoh\confrep{}{, or
they may require tedious reasoning about $\lambda$-lifted functions or {\sf SKBCI}-style
combinators}.
An example is the conjecture $\exists f.\>\forall x\> y.\; f\>x\;y \eq
\cst{g}\;y\;x$, whose proof requires taking $\lambda x\> y.\; \cst{g}\;y\;x$
as the witness for~$f$.

@@

  \begin{sloppypar}
  Traditionally, most higher-order provers were designed from the ground up
  to target higher-order logic. Two
  exceptions are Otter-$\lambda$ by Beeson \cite{mb-04-lam-logic} and
  Zipperposition by Cruanes et al.\ %, Vukmirovi\'c, and colleagues
  \cite{sc-15-simon-phd,sc-supind-17}. \hbox{Otter-$\lambda$}
  adds $\lambda$-terms and second-order unification to
  the superposition prover Otter\begin{rep} \cite{mcc-03-otter}\end{rep}.
  % The approach is pragmatic. , with little emphasis on completeness.
  Zipperposition, also based on superposition, was extended
  to Bool\-e\-an-free higher-order logic by Bentkamp et al.\
  \cite{bbtvw-21-sup-lam}. Its performance is a far cry from E's, but
  it is easier to modify.
  Vukmirovi\'c et al.\ also used it to test and evaluate higher-order unification
  procedures \cite{vbn-21-unif} and Boolean reasoning
  \cite{vn-20-bools}.
  Zipperposition now includes Ehoh as a backend in a cooperative
  architecture.
  Finally, there is work by the developers of Vampire
  % br-20-full-sup-w-combs
  \cite{br-19-restricted-unif} and of the SMT solvers CVC4 and veriT
  \cite{brotb-19-ho-smt} to extend their provers to higher-order logic.
  \end{sloppypar}

### 2022-03-11 ho-techniques/ho-techniques.tex
intro

>>>>

\authors{
    Joint work with\\
    Alexander Bentkamp,
    Jasmin Blanchette,
    Simon Cruanes,
    Visa Nummelin, and Sophie Tourret
}

\blfootnote{In this work I was the main designer of all presented techniques,
with the exception of inference streams which were designed by Sophie Tourret
and Alexander Bentkamp. Alexander Bentkamp and Jasmin Blanchette also discussed many
of the techniques with me and suggested important updates. Visa Nummelin worked on
implementation of FOOL preprocessing. Simon Cruanes is the original developer
of Zipperposition and provided us with invaluable knowledge.
}

\begin{abstract}
    Superposition is among the most successful calculi for first-order logic. Its
    extension to higher-order logic introduces new challenges such as infinitely
    branching inference rules, new possibilities such as reasoning about
    Booleans, and the need to curb the explosion of specific higher-order rules. We
    describe techniques that address these issues and extensively evaluate their
    implementation in the Zipperposition theorem prover. Largely thanks to their use,
    Zipperposition won the higher-order division of the CASC competition in 2020.  
  \end{abstract}
\newpage

\section{Introduction}
\label{sec:ho-tech:intro}

% %In recent decades,
% Superposition-based first-order automatic theorem provers
% have emerged as useful reasoning tools. They dominate at the annual CASC
% \cite{gs-2016-casc} theorem prover competition, having always won the
% first-order theorem division. They are also used as backends to proof assistants
% \cite{ck-18-coqhammer,ku-15-holyhammer,pb-12-sh}, automatic
% higher-order theorem provers \cite{sb-21-leo3}, and software verifiers
% \cite{fp-13-why3}.

% The superposition calculus has only recently been extended
% to higher-order logic (more precisely, extensional simple type theory
% \cite{henkin-1950-completeness}), resulting in
% \emph{\lsup} \cite{bbtvw-21-sup-lam}, which we developed
% together with Waldmann, as well as \emph{combinatory superposition}
% \cite{br-20-full-sup-w-combs} by Bhayat and Reger. Although these two
% calculi do not support an interpreted Boolean type,
% they can be extended by ad hoc rules \cite{our-bool-paper} that support
% most of the Boolean reasoning necessary in practice.

% Both higher-order superposition calculi were designed to gracefully
% extend first-order reasoning. As most steps in higher-order
% proofs tend to be essentially first-order, extending the most successful first-order
% calculus to higher-order logic seemed worth trying.
% Our first attempt at testing this idea was in 2019:
% Zipperposition~1.5, based on \lsup, finished third
% in the higher-order theorem division of CASC-27 \cite{gs-19-casc27},
% 12~percentage points behind the winner, the tableau prover Satallax 3.4 \cite{cb-2013-satallax}.

The landscape of higher-order proving techniques based on extension of efficient
first-order ones has tremendously expanded in the late 2010s and early 2020s. As
mentioned is Sect.~\ref{sec:pre:ho-sup-calculi} we have implemented three
higher-order calculi---\lfsup{}, \lsup{}, and \osup{}---which extend first-order superposition in a graceful way.
Bhayat and Reger also gracefully extended superposition to higher-order logic using
\textsf{SKBCI} combinators \cite{br-20-full-sup-w-combs}. Significant progress has been
made on SMT front as well \cite{brotb-19-ho-smt}.

In 2019 we tested if the idea of gracefully extending first-order provers to
higher-order logic really improves state of the art for the first time. We implemented \lsup{} \cite{bbtvw-21-sup-lam} in
Zipperposition~1.5 with basic heuristics and rudimentary extensions of the
calculus to deal with Booleans. It finished third at that year's higher-order
division of CASC competition \cite{gs-19-casc27}, 12~percentage points behind the
winner, the tableau prover Satallax 3.4 \cite{cb-2013-satallax}.

Studying the competition results, we found that higher-order tableaux have some
advantages over higher-order superposition. To bridge the gap, we developed
techniques and heuristics that simulate tableaux in the context of saturation.
We implemented them in Zipperposition~2, which took part at the higher-order
division of CASC \cite{gs-21-cascj10} in 2020. This time, our prover won the
division, proving 84\% of the problems, a whole 20~percentage points ahead of
the runner-up, Satallax 3.4.

In this chapter, we describe the main techniques that explain this reversal of
fortunes. They cover most parts of a modern higher-order theorem prover, from
preprocessing to additional calculus rules to heuristics to backend integration.
Compared to the previous chapter, in which we discussed rules used to treat
Boolean terms, in this chapter we use a newer version of Zipperposition, based
on a newer calculus. Instead of \lsup{} augmented with ad hoc Boolean rules, we
work with {\osup} \cite{bbtv-21-full-ho-sup}, a principled extension of
superposition to full higher-order logic, including an interpreted Boolean type.

Many higher-order problems extensively use symbol definitions to simplify
their representation. We describe several ways to exploit the definitions,
%of which the most successful is
such as turning them into rewrite rules (Sect.~\ref{sec:ho-tech:preprocessing}).
%Interesting patterns can be observed in various higher-order problem encodings.
%We show how we can exploit these to simplify problems (Sect.~\ref{sec:ho-tech:preprocessing}).
%
By working on formulas rather than clauses, tableau techniques take a more
holistic view of a higher-order problem.
Through its support for delayed clausification and, more generally,
calculus-level formula manipulation, \osup{} enables us to
simulate most successful tableau techniques in a saturating prover
(Sect.~\ref{sec:ho-tech:formulas}). This calculus also supports \emph{Boolean selection
functions}, a mechanism that allows us to choose on which Boolean subterms
to perform inferences first.
We implemented some Boolean selection functions and
evaluated them (Sect.~\ref{sec:ho-tech:bool-select}).

The main drawback of both $\lambda$-superposition variants compared with
combinatory superposition is that they rely on rules that enumerate possibly
infinite sets of unifiers. We describe a mechanism that interleaves infinitely
branching inferences with the standard saturation process
(Sect.~\ref{sec:ho-tech:infinite-branching}). The prover retains the same
behavior as before on first-order problems, smoothly scaling with increasing
numbers of higher-order clauses.
%
We also propose some heuristics to curb the explosion induced by highly
prolific calculus rules (Sect.~\ref{sec:ho-tech:explosiveness}).

Using first-order backends to finish the proof is common practice in
higher-order reasoning. Since \osup{} coincides with standard
superposition on first-order clauses, invoking backends may
seem redundant; yet Zipperposition is nowhere as efficient as E
\cite{scv-19-e23} or Vampire \cite{lkav-13-vampire}, so invoking a more
efficient backend does make sense. We describe how to achieve a balance
between allowing native higher-order reasoning and
delegating reasoning to a backend (Sect.~\ref{sec:ho-tech:backends}).
%
Finally, we compare Zipperposition~2 with other provers on all monomorphic
higher-order TPTP benchmarks \cite{gs-17-tptp} to perform a more extensive
evaluation than at CASC (Sect.~\ref{sec:ho-tech:comparison}). Our evaluation
corroborates the competition results.

### 2022-03-11 introduction/introduction.tex
intro

  Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second phase described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in a short amount of time we decided to
  implement the calculus in Zipperposition. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It finished third, 1 percent point behind Leo-III
  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
  \cite{cb-12-satallax}.

@@

Various researchers also experimented with the idea of extending a first-order
prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second-order
unification and added $\lambda$-terms to one of the best superposition
provers at the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
\cite{sc-15-simon-phd} with arithmetic, induction, and rudimentary support for
higher-order logic. Bhayat and Reger implemented a complete superposition
calculus  for higher-order logic, based on combinators  in Vampire
\cite{br-20-full-sup-w-combs}. Authors of SMT (satisfiability modulo theories) solvers followed a similar path:
they extended cvc4~\cite{cbetal-11-cvc4} and veriT~\cite{bodf-09-veriT} to
support higher-order logic \cite{brotb-19-ho-smt}.

>>>>

  Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second phase described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in a short amount of time we decided to
  implement the calculus in Zipperposition \cite{sc-15-simon-phd,sc-supind-17}. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It finished third, 1 percent point behind Leo-III
  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
  \cite{cb-12-satallax}.

@@

Various researchers also experimented with the idea of extending a first-order
prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second-order
unification and added $\lambda$-terms to one of the best superposition
provers at the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
\cite{sc-15-simon-phd,sc-supind-17} with arithmetic, induction, and rudimentary support for
higher-order logic. Bhayat and Reger implemented a complete superposition
calculus  for higher-order logic, based on combinators  in Vampire
\cite{br-20-full-sup-w-combs}. Authors of SMT (satisfiability modulo theories) solvers followed a similar path:
they extended cvc4~\cite{cbetal-11-cvc4} and veriT~\cite{bodf-09-veriT} to
support higher-order logic \cite{brotb-19-ho-smt}.
